\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{array}

\lstset{
  basicstyle=\ttfamily\footnotesize,
  keywordstyle=\color{blue},
  commentstyle=\color{gray},
  stringstyle=\color{red},
  breaklines=true,
  frame=single,
  numbers=left,
  numberstyle=\tiny\color{gray},
  language=C
}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\begin{document}

\title{Terminal Snake: An Operating-Systems Case Study in\\
Custom Memory, I/O, Process Management, and\\
WebAssembly Cross-Platform Deployment}

\author{
  \IEEEauthorblockN{Navnit Naman}
  \IEEEauthorblockA{\textit{School of Technology}\\
  \textit{Rishihood University}\\
  Sonipat, India\\
  230085}
  \and
  \IEEEauthorblockN{Kanhaiya Kumar}
  \IEEEauthorblockA{\textit{School of Technology}\\
  \textit{Rishihood University}\\
  Sonipat, India\\
  230062}
}

\maketitle

% ─────────────────────────────────────────────────────────────────────────────
\begin{abstract}
This paper presents \textit{Terminal Snake}, a real-time game written in ISO
C99 that intentionally avoids the C standard library's memory, string, and
math subsystems. The project serves as a self-contained demonstration of core
operating-systems concepts: a best-fit heap allocator with bidirectional
coalescing, raw-mode terminal I/O via POSIX \texttt{termios}, ANSI/VT100
screen rendering, a cooperative game-loop scheduler, and structured
error-handling. All five OS-course evaluation domains are addressed. Beyond the
terminal build, the project was ported to the browser via Emscripten and
WebAssembly: the same game C source compiles to \texttt{.wasm} and runs inside
an xterm.js terminal in any modern browser, with procedural sound generated by
the Web Audio API and high-score persistence via \texttt{localStorage}. Two
distinct game modes (Classic and Wrap) are implemented. A consecutive-food
combo multiplier (up to $\times$8), a lives/respawn system, an obstacle
system, and adaptive difficulty scaling make the gameplay rich while remaining
fully implemented in hand-crafted C. Version history is maintained on GitHub
under \texttt{kanhaiyak23/Snake\_game\_}.
\end{abstract}

\begin{IEEEkeywords}
operating systems, memory management, process management, I/O management,
custom allocator, terminal game, WebAssembly, Emscripten, Web Audio API,
C99, POSIX, best-fit allocation, coalescing, cross-platform
\end{IEEEkeywords}

% ─────────────────────────────────────────────────────────────────────────────
\section{Introduction}

Most hobby games lean on \texttt{malloc}, \texttt{printf}-wrappers, and
framework event loops without considering the subsystems underneath. This
project inverts that approach: the constraint of \emph{no standard-library
memory, string, or math functions} forces every OS primitive to be made
explicit and auditable.

The five pillars addressed are:

\begin{itemize}
  \item \textbf{Memory Management} --- a 1\,MiB static pool with best-fit
        allocation and bidirectional free-block coalescing.
  \item \textbf{I/O Management} --- raw-mode keyboard capture via
        \texttt{termios} and ANSI/VT100 screen rendering; in the browser,
        replaced by JavaScript key-event injection and xterm.js output.
  \item \textbf{Process Management} --- a deterministic cooperative game loop
        with \texttt{usleep}-based frame pacing (terminal) or
        \texttt{emscripten\_set\_main\_loop} (browser) and level-dependent
        speed scaling.
  \item \textbf{Security / Error Handling} --- structured termination codes,
        double-free guards, out-of-bounds checks, and graceful OOM recovery.
  \item \textbf{File Systems / Persistence} --- high-score persistence via the
        browser's \texttt{localStorage} API; a plain-file equivalent is
        documented for the terminal build in future scope.
\end{itemize}

The codebase has two compilation targets that share all game logic but swap in
platform-specific I/O implementations through the build system, demonstrating
a \emph{hardware abstraction layer} design without conditional compilation
inside library headers.

% ─────────────────────────────────────────────────────────────────────────────
\section{System Architecture}

\subsection{Dual-Platform Module Layout}

The project is divided into three layers: shared custom libraries, shared game
logic, and platform-specific I/O drivers:

\begin{table}[h]
\centering
\caption{Source Module Overview}
\begin{tabular}{lp{5.0cm}}
\toprule
\textbf{Module} & \textbf{Responsibility} \\
\midrule
\multicolumn{2}{l}{\textit{Shared custom libraries (both targets)}} \\
\texttt{math.c}     & Arithmetic (mul, div, mod, abs, clamp), XorShift32 PRNG \\
\texttt{string.c}   & strlen, strcpy, strcmp, itoa, strtok replacements \\
\texttt{memory.c}   & 1\,MiB static pool, best-fit allocator, coalescing \\
\texttt{sound.h}    & Sound API declaration (platform-independent) \\
\midrule
\multicolumn{2}{l}{\textit{Shared game logic (both targets)}} \\
\texttt{snake.h/c}  & Game structs, movement, collision, scoring, lives, obstacles, wrap mode \\
\midrule
\multicolumn{2}{l}{\textit{Terminal-specific (native build)}} \\
\texttt{keyboard.c} & POSIX termios raw mode, escape-sequence state machine \\
\texttt{screen.c}   & ANSI/VT100 renderer, TIOCGWINSZ layout \\
\texttt{sound.c}    & Terminal beep stub (no audio hardware dependency) \\
\texttt{main.c}     & Start screen, countdown, game loop, game-over screen \\
\midrule
\multicolumn{2}{l}{\textit{Browser-specific (Emscripten/WASM build)}} \\
\texttt{keyboard\_wasm.c} & JS key-event → C ring buffer bridge \\
\texttt{screen\_wasm.c}   & Output buffer flushed to xterm.js via \texttt{HEAPU8} \\
\texttt{sound\_wasm.c}    & Procedural synthesis via Web Audio API \\
\texttt{main\_wasm.c}     & State machine, localStorage high score, Emscripten loop \\
\midrule
\texttt{test\_logic.c} & Non-interactive unit tests for library correctness \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Platform Abstraction Pattern}

All platform-specific functionality is isolated in leaf modules that implement
the same function signatures declared in shared headers. The \texttt{Makefile}
selects the correct set of \texttt{.c} files for each target:

\begin{itemize}
  \item \texttt{make} links \texttt{keyboard.c + screen.c + sound.c + main.c}
        to produce a native POSIX binary (\texttt{./snake}).
  \item \texttt{make wasm} invokes \texttt{emcc} and links
        \texttt{keyboard\_wasm.c + screen\_wasm.c + sound\_wasm.c +
        main\_wasm.c} to produce \texttt{docs/snake.wasm} and
        \texttt{docs/snake.js}.
\end{itemize}

No \texttt{\#ifdef} guards are needed in shared headers; the abstraction is
achieved purely through the linker, exactly as a real OS hardware abstraction
layer works.

\subsection{Constraint Rationale}

The project prohibits \texttt{<string.h>}, \texttt{<math.h>}, \texttt{malloc},
\texttt{free}, and \texttt{rand}. The only permitted standard-library symbols
are \texttt{printf}/\texttt{putchar} (terminal I/O) and \texttt{exit}
(process control). This constraint forces every memory allocation, random
number, and string operation to be implemented and understood from first
principles---precisely the pedagogical intent of an OS course project.

% ─────────────────────────────────────────────────────────────────────────────
\section{Memory Management}

\subsection{Pool Design}

A single \texttt{static char VRAM[1\,048\,576]} array constitutes the entire
heap. On \texttt{mem\_init()}, it is formatted as one free \texttt{BlockHeader}
spanning the whole array:

\begin{lstlisting}
typedef struct BlockHeader {
    int  size;              /* user payload bytes */
    int  used;              /* 1=allocated, 0=free */
    struct BlockHeader *next;
} BlockHeader;
\end{lstlisting}

All blocks are linked in address order, enabling $O(1)$ adjacency tests needed
for coalescing.

\subsection{Best-Fit Allocation}

On each \texttt{alloc(size)} call, the allocator:
\begin{enumerate}
  \item Rounds \texttt{size} up to the nearest 8-byte boundary (alignment).
  \item Scans the free list for the smallest block that fits (\emph{best-fit}
        policy), breaking early on an exact match.
  \item Splits the chosen block if the remainder exceeds
        \(\text{sizeof(BlockHeader)} + 8\) bytes.
  \item Returns \texttt{NULL} on exhaustion, triggering the OOM error path.
\end{enumerate}

Best-fit was chosen over first-fit because snake segments are all the same
size (\texttt{sizeof(Segment)} = 12 bytes). Under uniform allocation sizes,
best-fit leaves fewer unusable slivers \cite{knuth1997art}.

\subsection{Bidirectional Coalescing}

On \texttt{dealloc(ptr)}, the allocator:
\begin{enumerate}
  \item Validates the pointer is within \texttt{VRAM} (bounds guard).
  \item Checks \texttt{blk->used} to prevent double-free corruption.
  \item Marks the block free and coalesces \emph{rightward} (absorbs
        consecutive free successors in a \texttt{while} loop).
  \item Finds the predecessor by address-order traversal and coalesces
        \emph{leftward} if the predecessor is free and adjacent.
\end{enumerate}

This two-phase coalescing prevents fragmentation even under the
alloc-one-grow/dealloc-one-shrink pattern of every game tick.

\subsection{Per-Tick Allocation Pattern}

The snake body is a singly-linked list of \texttt{Segment} nodes, each
allocated from the pool. Every tick:
\begin{itemize}
  \item \texttt{alloc(sizeof(Segment))} creates a new head at the computed
        next position.
  \item If no food was eaten, \texttt{dealloc(tail)} removes the oldest
        segment, keeping pool usage bounded at exactly
        \texttt{length $\times$ sizeof(Segment)} bytes.
  \item On food consumption the tail is \emph{not} freed, growing the snake
        by one segment permanently.
\end{itemize}

\subsection{Live Memory HUD}

\texttt{mem\_used()} walks the block list and sums used payload bytes.
This value is displayed every frame in the HUD, making heap behaviour
observable during gameplay.

% ─────────────────────────────────────────────────────────────────────────────
\section{I/O Management}

\subsection{Terminal: Keyboard Input}

\texttt{keyboard\_init()} transitions the terminal into raw, non-canonical,
non-echoing mode via POSIX \texttt{termios}:

\begin{lstlisting}
new_tio.c_lflag &= ~(ICANON | ECHO);
new_tio.c_cc[VMIN]  = 0;  /* non-blocking */
new_tio.c_cc[VTIME] = 0;
tcsetattr(STDIN_FILENO, TCSANOW, &new_tio);
fcntl(STDIN_FILENO, F_SETFL, flags | O_NONBLOCK);
\end{lstlisting}

Setting \texttt{VMIN=0, VTIME=0} with \texttt{O\_NONBLOCK} makes
\texttt{read()} return immediately whether or not a character is waiting.
This is necessary for a real-time game loop that cannot block on user input.

\subsubsection{Escape-Sequence State Machine}

Arrow keys arrive as three-byte ANSI sequences (\texttt{ESC [ A/B/C/D} for
up/down/right/left). \texttt{keyPressed()} implements a four-state machine
(\texttt{esc\_state} $\in \{0, 1, 2, 3\}$) to reassemble these sequences
across successive \texttt{read()} calls and map them to the WASD key codes
used by game logic. Stale buffered input is drained between screens to
prevent bleed-through.

\subsubsection{Terminal Restoration}

\texttt{keyboard\_restore()} removes \texttt{O\_NONBLOCK} and calls
\texttt{tcsetattr} with the saved \texttt{old\_tio} snapshot, guaranteeing
the terminal is left in a working state on both normal exit and error paths.

\subsection{Browser: Keyboard Bridge}

In the browser build, \texttt{keyboard\_wasm.c} exports a C function
\texttt{wasm\_push\_key(int code)} via \texttt{EMSCRIPTEN\_KEEPALIVE}. The
HTML page registers a \texttt{keydown} event listener that maps browser key
codes (e.g.\ \texttt{ArrowUp}, \texttt{KeyW}) to ASCII characters and calls
\texttt{wasm\_push\_key} via \texttt{Module.ccall}. The key is stored in a
small ring buffer; \texttt{keyPressed()} dequeues one character per tick,
replicating the drain semantics of the termios implementation exactly.

\subsection{Terminal: Screen Output}

\texttt{screen.c} is a thin wrapper around ANSI/VT100 escape sequences:

\begin{itemize}
  \item \textbf{Alternate screen buffer} (\texttt{\textbackslash{}033[?1049h/l}):
        isolates game rendering from the user's shell history.
  \item \textbf{Cursor control} (\texttt{\textbackslash{}033[row;colH}):
        positions output anywhere on the terminal grid.
  \item \textbf{SGR color codes} (e.g., \texttt{\textbackslash{}033[92m} for
        bright green): style the snake, food, obstacles, and HUD.
  \item \textbf{Unicode box-drawing} (\texttt{╔═╗║╚╝}): rendered as raw UTF-8
        bytes via \texttt{screen\_putstr}, replacing the earlier ASCII border.
\end{itemize}

\texttt{screen\_update\_layout()} queries the terminal dimensions via
\texttt{ioctl(TIOCGWINSZ)} on every frame, allowing the game field to scale
dynamically. All game code accesses field geometry through accessor functions
(\texttt{screen\_field\_x/y/w/h()}) rather than cached constants.

\subsection{Browser: Screen Output Buffer}

\texttt{screen\_wasm.c} maintains a 65\,536-byte character buffer
\texttt{obuf[]} in C memory. Every \texttt{screen\_putchar} appends to this
buffer. On \texttt{screen\_flush()}, the entire buffer is transferred to
JavaScript as a \texttt{Uint8Array} via Emscripten's \texttt{HEAPU8.slice()}
view and passed to xterm.js's \texttt{term.write()} method. This batches all
rendering into a single JS call per frame, minimising JS/WASM boundary
crossings for performance.

% ─────────────────────────────────────────────────────────────────────────────
\section{WebAssembly Browser Build}
\label{sec:wasm}

\subsection{Emscripten Compilation}

The browser target is compiled with \texttt{emcc} (Emscripten C compiler),
which translates C99 source to WebAssembly bytecode. The output is a
\texttt{.wasm} binary loaded by a companion \texttt{.js} glue file. Both
files are placed in \texttt{docs/} and served directly from GitHub Pages,
making the game playable at any browser without installation.

Key Emscripten-specific features used:

\begin{itemize}
  \item \texttt{EM\_ASM\{...\}} --- inline JavaScript executed synchronously
        from C, used for all Web Audio API calls and localStorage access.
  \item \texttt{EM\_ASM\_INT\{...\}} --- inline JavaScript that returns an
        integer to C, used to read the stored high score from
        \texttt{localStorage}.
  \item \texttt{EMSCRIPTEN\_KEEPALIVE} --- prevents the C linker from
        dead-stripping functions that are called only from JavaScript (e.g.\
        \texttt{wasm\_push\_key}).
  \item \texttt{emscripten\_set\_main\_loop(tick, 10, 1)} --- registers the
        game tick function with the browser's event loop at 10\,Hz base rate,
        replacing the native \texttt{while(1) + usleep} pattern. The actual
        frame rate is adjusted per-tick via
        \texttt{emscripten\_set\_main\_loop\_timing}.
\end{itemize}

\subsection{State Machine Architecture}

\texttt{main\_wasm.c} implements a four-state finite state machine that
drives all game screens:

\begin{table}[h]
\centering
\caption{Browser Build State Machine}
\begin{tabular}{lll}
\toprule
\textbf{State} & \textbf{Screen} & \textbf{Transition} \\
\midrule
\texttt{STATE\_START}   & Splash / title screen  & SPACE/ENTER $\to$ MENU \\
\texttt{STATE\_MENU}    & Mode selector          & SPACE/ENTER $\to$ PLAYING \\
\texttt{STATE\_PLAYING} & Game field             & game\_over $\to$ DEAD \\
\texttt{STATE\_DEAD}    & Game-over screen       & SPACE $\to$ MENU, Q $\to$ START \\
\bottomrule
\end{tabular}
\end{table}

The splash screen displays the ASCII art title, keyboard controls, and the
combo system hint. The menu lets the player navigate between Classic Mode and
Wrap Mode using W/S or arrow keys before confirming with SPACE/ENTER.

\subsection{High-Score Persistence via \texttt{localStorage}}

The browser build persists the session high score using the browser's
\texttt{localStorage} key--value store:

\begin{lstlisting}[language=C]
static void hs_load(void) {
    high_score = EM_ASM_INT({
        return parseInt(
            localStorage.getItem('snakeHS') || '0', 10);
    });
}
static void hs_save(void) {
    EM_ASM({
        try { localStorage.setItem('snakeHS', $0); }
        catch(e) {}
    }, high_score);
}
\end{lstlisting}

\texttt{hs\_load()} is called once at startup; \texttt{hs\_save()} is called
only when a new high score is achieved, minimising storage writes. The
\texttt{try/catch} in \texttt{hs\_save} handles private-browsing environments
where \texttt{localStorage} throws a \texttt{SecurityError}. The stored value
is presented on the game-over screen and highlighted with a ``New Personal
Best!'' message when broken.

% ─────────────────────────────────────────────────────────────────────────────
\section{Procedural Sound System}
\label{sec:sound}

\subsection{Design Rationale}

The terminal build has no reliable, portable audio hardware path, so
\texttt{sound.c} provides no-op stubs. The browser build implements a full
procedural sound engine in \texttt{sound\_wasm.c} using the Web Audio API
(\texttt{AudioContext}), with no audio files or external assets.

The \texttt{AudioContext} is lazily created on first use and cached in
\texttt{window.\_audioCtx}. If the context is suspended (browser autoplay
policy), \texttt{ctx.resume()} is called before scheduling notes. Three
internal note schedulers handle different wave types: \texttt{sched\_sine},
\texttt{sched\_square}, and \texttt{sched\_saw}.

\subsection{Implemented Sounds}

\begin{table}[h]
\centering
\caption{Procedural Sound Events}
\begin{tabular}{lp{4.8cm}}
\toprule
\textbf{Function} & \textbf{Description} \\
\midrule
\texttt{sound\_eat(combo)} & Two-phase sine ``coin blip''. Base pitch starts at 660\,Hz and rises 40\,Hz per combo level (capped at 920\,Hz). A perfect-fifth (3:2 ratio) upper note follows 35\,ms later. \\
\texttt{sound\_die()} & Three descending sawtooth tones: A4 (440\,Hz) $\to$ E4 (330\,Hz) $\to$ A3 (220\,Hz), spaced 95\,ms apart. Sawtooth timbre gives a harsh death-klaxon character. \\
\texttt{sound\_levelup()} & Five-note ascending square-wave fanfare: G4$\to$B4$\to$D5$\to$G5$\to$D6, spaced 85\,ms apart. Square wave produces the retro 8-bit aesthetic. \\
\texttt{sound\_game\_start()} & C major chord bloom: C4, G4, C5, G5 staggered by 60\,ms each, using sine waves for a clean, warm attack. \\
\texttt{sound\_wrap()} & FM synthesis portal wobble: an 8\,Hz LFO (modulator) with 180\,Hz depth drives a carrier sweeping from 420\,Hz to 200\,Hz over 200\,ms, producing a warp-pipe warble. \\
\bottomrule
\end{tabular}
\end{table}

\subsection{FM Synthesis Implementation}

\texttt{sound\_wrap()} demonstrates FM (frequency modulation) synthesis, a
technique normally requiring floating-point DSP:

\begin{lstlisting}[language=C]
var mod = ctx.createOscillator();   /* 8 Hz LFO */
var modGain = ctx.createGain();     /* depth: 180 Hz */
var car = ctx.createOscillator();   /* carrier: 420->200 Hz */
mod.connect(modGain);
modGain.connect(car.frequency);     /* LFO drives carrier */
\end{lstlisting}

The modulator oscillator's output (scaled by the gain node) is routed
directly into the carrier oscillator's \texttt{frequency} AudioParam,
creating periodic pitch variation. The carrier simultaneously sweeps
downward via \texttt{exponentialRampToValueAtTime}, combining the two
effects into a single warbling tone.

% ─────────────────────────────────────────────────────────────────────────────
\section{Process Management}

\subsection{Terminal: Game Loop Architecture}

The terminal game runs under a cooperative, single-process loop in
\texttt{main.c}:

\begin{enumerate}
  \item \texttt{keyPressed()} --- non-blocking poll for user input.
  \item \texttt{game\_handle\_input()} --- apply direction change or pause/quit.
  \item \texttt{game\_update()} --- advance physics by one tick.
  \item \texttt{game\_render()} --- redraw the entire frame.
  \item \texttt{usleep(game\_speed\_delay())} --- yield CPU for the remainder
        of the frame budget.
\end{enumerate}

This is a textbook \emph{fixed-timestep cooperative loop} \cite{gregory2018game}.
There is no preemptive multitasking; the process voluntarily yields via
\texttt{usleep}, which on macOS/Linux maps to \texttt{nanosleep} at the kernel
level.

\subsection{Browser: Emscripten Event Loop}

In the browser build, \texttt{usleep} and \texttt{while(1)} are replaced by:

\begin{lstlisting}[language=C]
emscripten_set_main_loop(tick, 10, 1);
\end{lstlisting}

This registers \texttt{tick()} with the browser's \texttt{requestAnimationFrame}
scheduler at a nominal 10 calls/second. The actual rate is adjusted per-tick:

\begin{lstlisting}[language=C]
emscripten_set_main_loop_timing(
    EM_TIMING_SETTIMEOUT,
    game_speed_delay(&gs) / 1000);  /* µs -> ms */
\end{lstlisting}

This converts the same microsecond delay value used by the terminal build into
milliseconds for the browser's \texttt{setTimeout}-based scheduler, giving
both targets identical frame pacing without any game-logic changes.

\subsection{Frame-Rate Scaling}

\texttt{game\_speed\_delay()} computes the inter-frame sleep duration as a
function of level:

\[
  \text{delay}(\ell) = \text{clamp}(180{,}000 - 15{,}000 \times (\ell - 1),\; 50{,}000,\; 180{,}000) \;\mu\text{s}
\]

Level~1 yields a 180\,ms frame period (5.6 ticks/s), and level~10 yields
50\,ms (20 ticks/s). All arithmetic uses \texttt{my\_mul} and \texttt{my\_clamp}
from the custom math library.

\subsection{Level Progression}

Level advances every 50 score points:

\[
  \ell = \min\!\left(\left\lfloor \frac{\text{score}}{50} \right\rfloor + 1,\; 10\right)
\]

Because the score multiplier (Section~\ref{sec:multiplier}) accelerates point
gain, high-skill play naturally reaches maximum speed faster.

\subsection{Startup Sequence}

The ordered startup sequence enforces a safe initialisation contract:
\texttt{mem\_init} $\to$ \texttt{keyboard\_init} $\to$ \texttt{screen\_init}
$\to$ \texttt{my\_srand} $\to$ \texttt{hs\_load} $\to$ game loop $\to$
\texttt{game\_free} $\to$ \texttt{screen\_cleanup} $\to$
\texttt{keyboard\_restore}. Resource acquisition and release are strictly
paired, preventing leaks across play-again sessions.

% ─────────────────────────────────────────────────────────────────────────────
\section{Security and Error Handling}
\label{sec:security}

\subsection{Structured Termination Codes}

Game-ending conditions are represented as symbolic constants rather than
magic integers:

\begin{table}[h]
\centering
\caption{Game-End Reason Codes}
\begin{tabular}{ll}
\toprule
\textbf{Constant} & \textbf{Trigger} \\
\midrule
\texttt{GAME\_END\_NONE}      & Game still running \\
\texttt{GAME\_END\_COLLISION} & Wall, obstacle, or self-collision \\
\texttt{GAME\_END\_OOM}       & \texttt{alloc()} returned \texttt{NULL} \\
\texttt{GAME\_END\_QUIT}      & User pressed \texttt{Q} \\
\texttt{GAME\_END\_INTERNAL}  & Linked-list invariant violated \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Memory Safety}

\begin{itemize}
  \item \textbf{Bounds guard:} \texttt{dealloc()} checks the pointer lies
        within \texttt{VRAM}; out-of-range pointers are silently ignored.
  \item \textbf{Double-free guard:} the \texttt{used} flag is checked before
        marking a block free; a second \texttt{dealloc} on the same pointer is
        a no-op.
  \item \textbf{OOM propagation:} \texttt{alloc()} returning \texttt{NULL} is
        checked at every call site; the game transitions to
        \texttt{GAME\_END\_OOM} rather than dereferencing a null pointer.
  \item \textbf{Invariant assertion:} if the tail-search loop cannot locate
        the tail pointer, the tick is aborted and \texttt{GAME\_END\_INTERNAL}
        is set, preventing use-after-free.
\end{itemize}

\subsection{Input Validation}

\begin{itemize}
  \item \textbf{Boundary check:} before placing the new head,
        \texttt{my\_in\_bounds()} verifies the position in Classic mode; in
        Wrap mode, modular arithmetic is applied instead of a bounds check.
  \item \textbf{Anti-reversal:} \texttt{game\_handle\_input()} blocks the
        180-degree direction change by checking the current velocity before
        updating \texttt{dx}/\texttt{dy}.
  \item \textbf{Input draining:} the keyboard buffer is flushed between menu
        screens, preventing stale escape sequences from triggering unintended
        game actions.
\end{itemize}

\subsection{Terminal Safety}

Both \texttt{keyboard\_restore()} and \texttt{screen\_cleanup()} are always
called on the exit path, ensuring the terminal is never left in raw mode or
on the alternate screen buffer.

% ─────────────────────────────────────────────────────────────────────────────
\section{Gameplay Features}

\subsection{Two Game Modes}
\label{sec:modes}

The menu screen offers two distinct modes selectable before each game:

\begin{itemize}
  \item \textbf{Classic Mode} --- hitting any wall or obstacle costs a life
        and triggers respawn. Wall collision is detected with
        \texttt{my\_in\_bounds(nx, ny, field\_w, field\_h)}.
  \item \textbf{Wrap Mode} --- the snake exits one edge and reappears on the
        opposite edge, using modular arithmetic:
\end{itemize}

\begin{lstlisting}[language=C]
/* Wrap mode: teleport through walls */
nx = my_mod(nx + field_w, field_w);
ny = my_mod(ny + field_h, field_h);
\end{lstlisting}

In Wrap mode, \texttt{sound\_wrap()} is triggered on each wall crossing to
give audio feedback for the teleportation. Self-collision and obstacle
collision still end a life in Wrap mode.

\subsection{Core Tick Loop}

Each call to \texttt{game\_update()} performs, in order:
(1) layout refresh, (2) out-of-bounds food relocation, (3) head position
computation, (4) boundary/wrap handling, (5) self-collision check,
(6) obstacle collision check, (7) food collision detection, (8) new head
allocation, (9) tail removal or retention, (10) score/level update.
This strict ordering prevents read-after-write hazards within a single tick.

\subsection{Score Multiplier (Combo System)}
\label{sec:multiplier}

A consecutive-food combo system rewards sustained play without dying. The
\texttt{combo} counter in \texttt{GameState} increments on each food eaten and
resets to zero on any life loss. The score awarded per food is:

\[
  \Delta\text{score} = 10 \times \ell \times \min(\text{combo},\; M_{\max})
\]

where $\ell$ is the current level and $M_{\max} = 8$ is the multiplier cap
(\texttt{MAX\_COMBO\_MULT}). The HUD highlights the multiplier in bold yellow
whenever it exceeds $1\times$, giving the player real-time streak feedback.
The eat-sound pitch also rises with combo level, providing audio confirmation
of the streak.

\subsection{Lives System}

The player begins with \texttt{INITIAL\_LIVES} = 3 lives.
\texttt{lose\_life\_and\_respawn()} decrements the counter, resets the combo,
clears all segment memory, rebuilds the snake at field centre, and repositions
all food items. When lives reach zero, the game ends with
\texttt{GAME\_END\_COLLISION}. Session state (score, high score, level)
persists across respawns within one game.

\subsection{Obstacle System}

Each game session places a random number of obstacles on the field. Obstacles
are statically allocated as an array \texttt{obstacles[MAX\_OBSTACLES]} (with
\texttt{MAX\_OBSTACLES} = 10) inside \texttt{GameState}---no dynamic
allocation is used. Placement uses the same rejection-sampling loop as food
placement: a candidate position is rejected if it overlaps the snake, any food
item, any existing obstacle, or the center spawn area. In Classic mode,
hitting an obstacle costs a life; in Wrap mode, the same rule applies since
wall-wrapping does not protect against obstacles. Obstacles are drawn in a
distinct colour to differentiate them from the snake body and food items.

\subsection{Multi-Food Spawning}

Up to \texttt{MAX\_FOODS} = 3 food items are active simultaneously.
\texttt{place\_food()} uses a rejection-sampling loop: a candidate position
is drawn from \texttt{my\_rand\_range()} and rejected if it overlaps any snake
segment, obstacle, or any other active food item. Since the snake is never
allowed to fill the field (collision terminates the session first), the loop
always terminates in finite expected time.

% ─────────────────────────────────────────────────────────────────────────────
\section{File System and Persistence}

\subsection{Browser: \texttt{localStorage} High Score}

The browser build fully implements high-score persistence using
\texttt{localStorage}, the browser's built-in key--value store that persists
data across page refreshes and sessions for the same origin. The key used is
\texttt{'snakeHS'}. The score is loaded once at startup and saved only when
beaten, making the storage interaction lightweight.

\subsection{Terminal: Deferred File I/O}

The terminal build currently holds all game state in memory with no disk
persistence. A minimal persistence layer would require:
\begin{itemize}
  \item A fixed-size binary struct serialised to \texttt{\textasciitilde/.snake\_score}.
  \item \texttt{open()}/\texttt{read()}/\texttt{write()}/\texttt{close()} calls
        (permitted as a POSIX hardware-abstraction exception).
  \item Atomic write-and-rename (\texttt{rename()} over a temporary file) to
        avoid partial writes on power loss.
\end{itemize}

This is documented as the highest-priority future item (Section~\ref{sec:future}).

% ─────────────────────────────────────────────────────────────────────────────
\section{Custom Math Library}

\subsection{Multiplication: Shift-and-Add}

\texttt{my\_mul(a, b)} uses the Russian-peasant (binary) method: the larger
magnitude is doubled in each step while the smaller is halved; the larger is
accumulated when the LSB of the smaller is 1. Time complexity is
$O(\log\min(|a|,|b|))$ bit operations; space is $O(1)$.

\subsection{Division: Recursive Doubling}

\texttt{divide\_positive\_ll(a, b)} halves the problem by doubling the
divisor until it exceeds the dividend, then unwinds the recursion accumulating
the quotient in binary. Recursion depth is $O(\log(a/b))$. \texttt{my\_div}
wraps this with sign handling (truncation toward zero).

\subsection{XorShift32 PRNG}

The random number generator uses the XorShift32 algorithm \cite{marsaglia2003xorshift}:

\begin{lstlisting}
rng_state ^= rng_state << 13;
rng_state ^= rng_state >> 17;
rng_state ^= rng_state << 5;
\end{lstlisting}

This produces a full-period sequence of $2^{32}-1$ values with no multiply
in the hot path. It is seeded from \texttt{time(NULL)} at startup. Both the
terminal and browser builds use the same PRNG; the browser seeds via
\texttt{time()} which Emscripten maps to \texttt{Date.now()}.

% ─────────────────────────────────────────────────────────────────────────────
\section{Results and Metrics}

\begin{table}[h]
\centering
\caption{Implementation Metrics}
\begin{tabular}{lr}
\toprule
\textbf{Metric} & \textbf{Value} \\
\midrule
Total source lines (excl.\ blank/comments) & $\approx 1{,}300$ \\
Shared game + library lines                 & $\approx 870$ \\
Browser-specific (wasm) lines               & $\approx 430$ \\
Heap pool size                              & 1\,048\,576 bytes (1\,MiB) \\
Segment size (allocated unit)              & 12 bytes \\
Maximum live segments (level 10, length 200) & $\approx 200$ \\
Peak heap usage at max length              & $\approx 4{,}800$ bytes (0.46\%) \\
Frame period range                          & 50\,ms -- 180\,ms \\
Max simultaneous food items                 & 3 \\
Max obstacles per session                   & 10 \\
Combo multiplier cap                        & 8$\times$ \\
Procedural sound events                     & 5 \\
Game modes                                  & 2 (Classic, Wrap) \\
Unit tests passing                          & 22 / 22 \\
\bottomrule
\end{tabular}
\end{table}

\noindent The allocator is empirically fragmentation-free for the snake
workload: each tick allocates one segment and (on non-eat frames) frees one
segment of identical size, so the pool converges to a near-steady-state after
a few ticks.

% ─────────────────────────────────────────────────────────────────────────────
\section{Future Scope}
\label{sec:future}

\begin{enumerate}
  \item \textbf{Terminal High-Score Persistence (File System):}
        Serialise the high score to a dot-file using POSIX
        \texttt{open/write/rename} with atomic replace semantics, matching
        the localStorage behaviour already implemented in the browser build.
  \item \textbf{Slab Allocator:}
        Since all game allocations are the same size (\texttt{sizeof(Segment)}),
        a slab/pool allocator \cite{bonwick1994slab} would reduce per-allocation
        overhead from $O(n)$ (best-fit scan) to $O(1)$ (free-list pop).
  \item \textbf{Configurable Difficulty:}
        Read a plain-text \texttt{.snakerc} configuration file at startup to
        set starting lives, field size, food count, and obstacle count.
  \item \textbf{Replay Recording:}
        Log the RNG seed and input stream to a file; replay by re-seeding
        and feeding the recorded keystrokes, enabling deterministic replay
        without storing frame buffers.
  \item \textbf{Multiplayer over Sockets:}
        Two game instances could share state via a UNIX-domain socket,
        exercising IPC and synchronisation primitives.
  \item \textbf{Signal Handling:}
        Install a \texttt{SIGWINCH} handler to respond to terminal resize
        events asynchronously rather than polling \texttt{TIOCGWINSZ} every
        frame, reducing unnecessary system calls.
  \item \textbf{IndexedDB Leaderboard:}
        Extend the browser build's high-score system to a full top-5 table
        using the browser's \texttt{IndexedDB} API, a structured alternative
        to \texttt{localStorage} for larger datasets.
\end{enumerate}

% ─────────────────────────────────────────────────────────────────────────────
\section{Version History and Collaboration}

All source code, commit history, and issue tracking are maintained at
\texttt{github.com/kanhaiyak23/Snake\_game\_}. Contributions from both
authors are individually attributable through commit authorship.
Key milestones recorded in the history include:

\begin{itemize}
  \item Initial snake movement and collision (\texttt{e1d954b})
  \item Lives system with respawn (\texttt{fb0e676})
  \item Multi-food spawning and HUD (\texttt{915acd4})
  \item Input-parsing improvements and layout accessors (\texttt{205a8bc})
  \item Score multiplier (combo system) and Unicode border (\texttt{766dafc})
  \item WebAssembly build: Emscripten port with xterm.js renderer (\texttt{eaa3cdd})
  \item Procedural sound engine via Web Audio API (\texttt{e6ce5e2})
  \item Wrap mode, obstacle system, and browser state machine (\texttt{e6ce5e2})
\end{itemize}

% ─────────────────────────────────────────────────────────────────────────────
\section{Conclusion}

Terminal Snake demonstrates that a complete, dual-platform game can be built
atop hand-crafted OS primitives without sacrificing correctness or safety.
The best-fit allocator with bidirectional coalescing correctly manages the
alloc/dealloc pattern of the game loop under all tested conditions. The
\texttt{termios} raw-mode keyboard driver and ANSI/VT100 renderer provide
responsive I/O in the terminal with deterministic restoration. The Emscripten
port proves that the platform-abstraction design---shared game logic, swappable
I/O drivers---scales cleanly from a native binary to a WebAssembly module
without modifying any game code. The procedural sound system, wrap-mode
teleportation, obstacle placement, combo multiplier, and \texttt{localStorage}
persistence collectively expand the project well beyond a minimal OS
demonstration into a feature-complete, browser-playable arcade game. Every
new feature is grounded in an OS concept: the sound engine exercises
asynchronous scheduling; wrap mode applies modular arithmetic; persistence
exercises the file-system module; and the state machine formalises process
state transitions. The project provides a concrete, measurable reference
implementation for each concept covered in the Operating Systems curriculum.

% ─────────────────────────────────────────────────────────────────────────────
\begin{thebibliography}{00}

\bibitem{knuth1997art}
D.~E. Knuth, \textit{The Art of Computer Programming, Vol.~1: Fundamental
Algorithms}, 3rd ed. Reading, MA: Addison-Wesley, 1997, pp.~435--456.

\bibitem{gregory2018game}
J.~Gregory, \textit{Game Engine Architecture}, 3rd ed. Boca Raton, FL: CRC
Press, 2018, pp.~391--420.

\bibitem{bonwick1994slab}
J.~Bonwick, ``The slab allocator: An object-caching kernel memory allocator,''
in \textit{Proc. USENIX Summer Technical Conf.}, Boston, MA, Jun. 1994,
pp.~87--98.

\bibitem{marsaglia2003xorshift}
G.~Marsaglia, ``Xorshift RNGs,'' \textit{J. Statistical Software}, vol.~8,
no.~14, pp.~1--6, 2003.

\bibitem{silberschatz2018os}
A.~Silberschatz, P.~B. Galvin, and G.~Gagne, \textit{Operating System
Concepts}, 10th ed. Hoboken, NJ: Wiley, 2018.

\bibitem{kerrisk2010linux}
M.~Kerrisk, \textit{The Linux Programming Interface}. San Francisco, CA:
No Starch Press, 2010, pp.~1331--1360.

\bibitem{ieee2016posix}
IEEE and The Open Group, \textit{The Open Group Base Specifications Issue 7,
2018 Edition (POSIX.1-2017)}, IEEE Std 1003.1-2017, 2017.

\bibitem{emscripten2023}
A.~Zakai, ``Emscripten: An LLVM-to-JavaScript Compiler,'' in \textit{Proc.
ACM Int. Conf. Object Oriented Programming Systems Languages and Applications
(OOPSLA)}, 2011, pp.~301--312.

\bibitem{webaudio2023}
P.~Adenot and H.~Choi, \textit{Web Audio API}, W3C Recommendation, Jun.\ 2021.
[Online]. Available: \url{https://www.w3.org/TR/webaudio/}

\end{thebibliography}

\end{document}
